Finally, a functional class scoring method for enrichment analysis of Infinium DNA methylation data

### 2023-12-20 manuscript/manuscript.Rmd
writing

* DNA methylation is the most widely studied epigenetic mark, for its role in development and
disease.

* Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts.

* Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing.

* Infinium arrays typically have multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers, which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.

* Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
class scoring (FCS).

In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list consisting of genes measured in the assay.
The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
gene list of interest relative to the background.

Functional class scoring takes a different approach by ranking all detected genes by a differential
regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
set of genes has distribution of scores that is different to the null.
Currently there are no specialised approaches for the application of FCS to Infinium array data.

@@

@@

### Pathway-level differential methylation in hyperhomocysteinemia
(2) the association of elevated plasma homocysteine with methylation patterns.

@@

To compare the sensitivity of these methods with real data, sample sizes were randomly downsampled, analysed
as above and the significant pathways were compared to the results from the full
group of 37 patients.
This process was repeated 50 times and the results shown in (**Figure 4**).
Pathways that were identified in the smaller and full group were termed "consistent",
while pathays that were identified in the smaller but not in the full group were
termed "inconsistent".
LAM was able to detect 300 out of 358 consistent pathways with a sample size of 10,
while GSA detected just 4 of 21 at this sample size.
LAM identified more inconsistent pathways than GSA, but the proportion of
inconsistent findings was lower across all sample sizes.
At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and
60% for GSA.
This result suggests that LAM has superior sensitivity to detect differentially methylated
pathways in real cancer data.

![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.](../figs/fig4.png)

>>>>

* DNA methylation is the most widely studied epigenetic mark, for its role in development and
disease [@Robertson2005-tu].

* Many hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts [@Li2019-og].

* Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing [@Wei2021-qj].

* Infinium arrays typically have multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.

* Enrichment analysis comes in two popular types; over-representation analysis (ORA) and functional
class scoring (FCS) [@Khatri2012-qy].
A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].

In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list consisting of genes measured in the assay.
The test seeks to identify sets of genes (eg: ontologies, pathways) that are over-represented in the
gene list of interest relative to the background [@Tavazoie1999-ng].

FCS takes a different approach by ranking all detected genes by a differential
regulation score (eg: fold change, confident effect size) followed by a test to assess whether each
set of genes has distribution of scores that is different to the null [@Subramanian2005-ar].

The `gsameth()` function of the `missMethyl` package is the state-of-the-art methods for ORA of infinium methylation array
data as it addresses issues related to probes belonging to more than one gene and the fact that genes can have multiple
probes [@Maksimovic2021-ur].
As gsameth is an ORA method, results dependent strongly on the significance threshold used [@Ren2019-mr], and also
on the proportion of probes that meet this threshold.

FCS methods `ebGSEA` and `methylGSA` have been applied to EWAS data and are suggested to have better sensitivity to
determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
enrichment [@Dong2019-np].
The `methylGSA` package implements two different approaches.
Firstly, the `methylRRA` function which uses robust rank aggregation for detecting differentially methylated genes taking
into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside eachgene set [@Ren2019-mr].
These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
changes on genomic regulation.
Neither gsameth, ebGSEA nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
enrichment score is a useful surrogate for effect size.
Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
biases towards large gene sets that have non-specific biological functions.
This can lead to end users over-looking smaller pathways with more specific biological functions that have larger effect
sizes, as those p-values are relatively larger.

Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that reports enrichment
scores.

@@

### Pathway-level differential methylation in aging
In order to deminstrate the utility of the LAM method for 
(2) the association of elevated plasma homocysteine with methylation patterns.

@@

@@

To compare the sensitivity of these methods with real data, sample sizes were randomly downsampled, analysed
as above and the significant pathways were compared to the results from the full
group of 37 patients.
This process was repeated 50 times and the results shown in (**Figure 4**).
Pathways that were identified as statistically significant (FDR<0.05) in the smaller and full group were
termed "consistent", while pathways that were identified in the smaller but not in the full group were
termed "inconsistent".
LAM was able to detect 300 out of 358 consistent pathways with a sample size of 10,
while GSA detected just 4 of 21 at this sample size.
LAM identified more inconsistent pathways than GSA, but the proportion of
inconsistent findings was lower across all sample sizes.
At a sample size of 10, the proportion of inconsistent pathways was 10% for LAM and
60% for GSA.
This result suggests that LAM has superior sensitivity to detect differentially methylated
pathways in real cancer data.

![Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.
Upper panels show the number of significant pathways identified in the smaller sample that were consistent
with the full group for LAM (left) and GSA (right).
Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
with the full group.
Lower panels show the proportion of significant pathways identified that were inconsistent with the full
group.
Left panels correspond to LAM, and right panels to GSA methods.
](../figs/fig4.png)
